Verifying Fine-Grained and Lock-Free Binary Search Trees

### 2021-12-29 concurrency/Verification of Concurrent Binary Search Tree.tex
finished first-draft explanation of BST lock invariants and ghost state

As described in section~\ref{atomicity}, we connect the per-node locks of our C implementation with the global abstract state of $\treerep$ by associating each node's lock with a lock variant assertion $\ell \islock_\pi R$, where $R$ is parameterized by the abstract state of the node (range + possible contents). We will fix the share $\pi$ to 1/2 so that the range in a node's lock can be out of date with respect to its $\mathsf{public\_half}$ (in particular, it may have grown due to rearrangements further up the tree). We then need to choose a variant $R$ describing the resources held by the lock for a given abstract state. The C implementation of a node is:
\begin{lstlisting}[language = C,numbers = none]
typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
\end{lstlisting}
The node data of each \lstinline{tree} is wrapped in a \lstinline{tree_t} struct, and is only accessible when the lock in the \lstinline{lock} field is held. The lock variant for a node at location $p$ then needs to describe the contents of the \lstinline{t} field for each possible abstract state:
\begin{align*}
&\mathsf{match}\ c\ \mathsf{with}\\
&|\ \mathsf{None} \Rightarrow \texttt{t} \mapsto \texttt{NULL}\\
&|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r)
\end{align*}
Because our implementation is doing hand-over-hand locking, in the internal case, we also need to obtain the lock variant assertions for the child nodes, so the full lock variant is: %this isn't enough; we need to talk about selflock and shares
\begin{align*}
\mathsf{node\_inv}((l, r), c) \triangleq \ &\mathsf{match}\ c\ \mathsf{with}\\
&|\ \mathsf{None} \Rightarrow \texttt{t} \mapsto \texttt{NULL}\\
&|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow k \in (l, r) \land \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r) * p_l \lockvar \mathsf{node\_inv} * p_r \lockvar \mathsf{node\_inv}
\end{align*}
Because we are using lock variants, we do not need to know the ranges or contents of the child nodes: instead, the variant construction guarantees that they are consistent with the global ghost state.

%%edit from here
We combine pieces of information discussed above, \emph{recursive lock invariant} from \ref{safety} and \emph{ghost states}, to write the atomic specification. The private pre- and postcondition of the atomic specification is given by defining the predicate $\nodeboxrep(p,g,g\_root)$ that describes the concrete state of the BST along with the contribution part of the both kind of ghost states explained above.The root node $p$ is represented by ghost name $g\_root$ and overall ghost state is represented by ghost name $g$. The $\nodeboxrep$ predicate ties together $\mathsf{my\_half}$ part of per-node ghost state and $\mathsf{in\_tree}$ of over-all ghost state inside the $\mathsf{lock\_inv}$ assertion of each node's lock as follows:
\begin{align*}
    &\nodeboxrep_{\pi}(p,g,g\_root) \triangleq  p\mapsto\ (lock,tp)\ *\ \mathsf{lock\_inv_{\pi}}(lock, \mathsf{selflock}_{0.5}(lock,R))) \\&R  \triangleq\ \ghost{\mathsf{my\_half}(range,info)}{g\_root}\ *\ \ghost{\mathsf{in\_tree}\ (g\_root)}{g}\ *\ \exists\ pa\ pb\ ga\ gb,\ tp\mapsto\ (pa * pb)\ *\  \nodeboxrep_{0.5}(pa,g,ga) \\& *\ \nodeboxrep_{0.5}(pb,g,gb))\end{align*}

The public pre- and postcondition of the atomic specification is given by defining the predicate $\treerep(T,g)$ that describes the abstract state of the binary search tree $T$  in terms of collection of ghost states each describing the information about separate node in the tree.. The $\treerep$ predicate ties together $\mathsf{public\_half}$ part of per-node ghost states and $\mathsf{ghost\_ref}$ of over-all ghost state as follows: \begin{align*}&\mathsf{ghost\_tree\_rep}(tg,range, g, g\_current) \triangleq\ \exists\ k, v, ga, gb\ \in tg,\ghost{\mathsf{public\_half}\ (range, Some(k,v,ga,gb))}{g\_current}\\& *\ \mathsf{ghost\_tree\_rep}(left(tg),(left(range),k),g,ga)\ *\ \mathsf{ghost\_tree\_rep}(right(tg),(k, right(range)),g,gb) \end{align*}
\begin{align*}\treerep(T,g,g\_root) \triangleq \ \exists\ (tg:ghost\_tree),\ &\mathsf{pure\_tree}(tg)\ =\ T\ \land\ \mathsf{ghost\_tree\_rep}(tg,(-\infty, +\infty),g,g\_root)&\\ *\ \ghost{\mathsf{ghost\_ref}\ (\mathsf{find\_ghost\_set}(tg))}{g}\end{align*}
Where $tg$ is called \emph{ghost tree}, which can be existentially quantified, extend the node in pure tree $T$ with actual pointer and ghost names for child nodes.

The atomic specification for any binary search tree's operation \texttt{op} should then be
\begin{align*}\forall t.\ \langle \nodeboxrep(p,g,g_{\mathit{root}})\ |\ \treerep(t,g)\rangle\ \texttt{op}(p)\ \langle \nodeboxrep(p,g,g_{\mathit{root}})\ |\ \treerep(t', g)\rangle
\end{align*}
where $t$ and $t'$ are the abstract state of a tree before and after the function's execution. Now we are prepared to prove that each of our BST functions satisfies its specification.

>>>>

As described in section~\ref{atomicity}, we connect the per-node locks of our C implementation with the global abstract state of $\treerep$ by associating each node's lock with a lock variant assertion $\ell \islock_\pi R$, where $R$ is parameterized by the abstract state of the node (range + possible contents). We will fix the share $\pi$ to 1/2 so that the range in a node's lock can be out of date with respect to its $\mathsf{public\_half}$ (in particular, it may have grown due to rearrangements further up the tree). We then need to choose a variant $R$ describing the resources held by the lock for a given abstract state. The C implementation of a node is:
\begin{lstlisting}[language = C,numbers = none]
typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
\end{lstlisting}
The node data of each \lstinline{tree} is wrapped in a \lstinline{tree_t} struct, and is only accessible when the lock in the \lstinline{lock} field is held. The lock variant for a node at location $p$ then needs to describe the contents of the \lstinline{t} field for each possible abstract state:
\begin{align*}
&\mathsf{match}\ c\ \mathsf{with}\\
&|\ \mathsf{None} \Rightarrow \texttt{t} \mapsto \texttt{NULL}\\
&|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r)
\end{align*}
Because our implementation is doing hand-over-hand locking, in the internal case, we also need to obtain the lock variant assertions for the child nodes, so the full lock variant is: %this isn't enough; we need to talk about selflock and shares
\begin{align*}
\mathsf{node\_inv}((l, r), c) \triangleq \ &\mathsf{match}\ c\ \mathsf{with}\\
&|\ \mathsf{None} \Rightarrow \texttt{t} \mapsto \texttt{NULL}\\
&|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow k \in (l, r) \land \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r) * \\&\qquad p_l.\texttt{lock} \lockvar \mathsf{node\_inv} * p_r.\texttt{lock} \lockvar \mathsf{node\_inv}
\end{align*}
While we do not know the ranges or contents of the child nodes, the lock-variant construction guarantees that they are consistent with the $\mathsf{public\_half}$ assertions of the global ghost state.

Now we have everything we need to specify the tree operations:
\begin{mathpar}
\forall m.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ m\ k\ v)\rangle

\forall m.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{lookup}\ m\ k = v\rangle

\forall m.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{delete}\ m\ k)\rangle
\end{mathpar}
where $\treerep(m)$ asserts that global ghost state of the tree implements a map $m$, as described above, and $\nodeboxrep$ (the client's handle to the data structure) is simply a reference to the lock variant for the root node:
$$\nodeboxrep(b) \triangleq \exists p.\ b \mapsto p * p.\texttt{lock} \lockvar \mathsf{node\_inv}$$

Now we are prepared to prove that each of our BST functions satisfies its specification.
